Dual host-parasite transcriptomes of apicomplexan Eimeria falciformis and its natural mouse host

### 2016-07-14 figuresANDmanuscript/manuscriptBMCformat/manuscript_rnaseq.tex
MS reshuffled and folder structure changed: MS nw i figures folder

\usepackage{graphicx}
\graphicspath{ {../} }
%\def\includegraphic{}
%\def\includegraphics{}

%%% Put your definitions there:
\startlocaldefs
%\affi1=Department of Molecular Parasitology, Humboldt University, Berlin, Germany
%\affi2=Max Planck Institute for Biology of Ageing, Joseph-Stelzmann-Strasse 9B, 50931 Cologne, Germany
%\affi3=Department Reproduction Biology and Evolutionary Ecology, Leibniz Institute of Zoo and Wildlife Research, Forschungsverbund Berlin e.V., Berlin, Germany
\endlocaldefs

@@

\title{Dual host-parasite transcriptomes of apicomplexan \textit{Eimeria falciformis} and its natural mouse host}

@@

\author[
   addressref={aff1},                   % id's of addresses, e.g. {aff1,aff2}
   corref={aff1},                       % id of corresponding address, if any
   noteref={n1},                        % id's of article notes, if any
   email={emanuel.heitlinger@hu-berlin.de}   % email address
]{\inits{EH}\fnm{Emanuel} \snm{Heitlinger}}
\author[
   addressref={aff1},
   %email={john.RS.Smith@cambridge.co.uk}
]{\inits{TK}\fnm{Totta} \snm{Kasemo}}

@@

\address[id=aff1]{%                           % unique id
  \orgname{Department of Biology, Humboldt-Universitat zu Berlin}, % university, etc
  \street{Philippstr. 13, Haus 14},                     %
  %\postcode{}                                % post or zip code
  \city{Berlin},                              % city
  \cny{Germany}                                    % country
}
\address[id=aff2]{%
  \orgname{Marine Ecology Department, Institute of Marine Sciences Kiel},
  \street{D\"{u}sternbrooker Weg 20},
  \postcode{24105}
  \city{Kiel},
  \cny{Germany}
}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%\author[
%   addressref={affi1},                   % id's of addresses, e.g. {aff1,aff2}
%   corref={aff1},                       % id of corresponding address, if any
%  % noteref={n1},                        % id's of article notes, if any
%   email={kasemotq@hu-berlin.de}   % email address
%]{\inits{TK}\fnm{Totta} \snm{Kasemo}}
%\author[
%   addressref={affi1}                   % id's of addresses, e.g. {aff1,aff2}
%  % corref={aff1},                       % id of corresponding address, if any
%  % noteref={n1},                        % id's of article notes, if any
%  % email={}   % email address
%]{\inits{SS}\fnm{Simone} \snm{Spork}}
%%\author[
%   addressref={affi2},                   % id's of addresses, e.g. {aff1,aff2}
%   %corref={aff3},                       % id of corresponding address, if any
%   noteref={n2},                        % id's of article notes, if any
%   %email={emanuel.heitlinger@hu-berlin.de}   % email address
%]{\inits{CD}\fnm{Christoph} \snm{Dieterich}}
%\author[
%   addressref={affi1, affi3},                   % id's of addresses, e.g. {aff1,aff2}
%   %corref={aff3},                       % id of corresponding address, if any
%   noteref={n1},                        % id's of article notes, if any
%   email={emanuel.heitlinger@hu-berlin.de}   % email address
%]{\inits{EH}\fnm{Emanuel} \snm{Heitlinger}}

%\address[id=aff1]{%                           % unique id
%  \orgname{Department of Zoology, Cambridge}, % university, etc
%  \street{Waterloo Road},                     %
  %\postcode{}                                % post or zip code
%  \city{London},                              % city
%  \cny{UK}                                    % country
%}
%\address[id=affi1]{%
%  \orgname{Humboldt-Universitat zu Berlin, Department of Molecular Parasitology},
%  \street{Philipstr. 13, House 14},
%  \postcode{10115}
%  \city{Berlin},
%  \cny{Germany}
%}
%\address[id=affi2]{%  
%  \orgname{Max Planck Institute for Biology of Ageing},
%  \street{Joseph-Stelzmann-Strasse 9B},
%  \postcode{50931}
%  \city{Cologne},
%  \cny{Germany}
%}
%\address[id=affi3]{%  
%  \orgname{Department Reproduction Biology and Evolutionary Ecology, Leibniz Institute of Zoo and Wildlife Research},
%  \street{Alfred-Kowalke-Str. 17},
%  \postcode{10315}
%  \city{Berlin},
%  \cny{Germany}
%}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                          %%
%% Enter short notes here                   %%
%%                                          %%
%% Short notes will be after addresses      %%
%% on first page.                           %%
%%                                          %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{artnotes}
%\note{Sample of title note}     % note to the article
%\note[id=n1]{Equal contributor} % note, connected to author
\end{artnotes}

\end{fmbox}% comment this for two column layout

@@

\end{abstractbox}
%
%\end{fmbox}% uncomment this for twcolumn layout

@@

  \section*{Results}
Text for this section \ldots
\subsection*{GO term enrichments in heatmap clusters}
The annotations referred to here are inferred from orthologs in other \textit{Eimeria} spp. or
in \textit{T. gondii}

\subsubsection*{Data overview}
In total, xx reads were sequenced. For each sample, mouse read numbers were 10⁷ - 10⁸
Quality filters applied resulted in xx reads being removed from further analysis

@@

\subsection{RNA extraction}
Total RNA was isolated from infected epithelial cells, sporozoites and sporulated oocysts 
using Trizol according to the manufacturer’s protocol (Invitrogen). High quality  \emph{what is the meaning of 'high quality' here?}
RNA was used to produce an mRNA library using the Illumina’s TruSeq RNA Sample Preparation guide.
\emph{stolen from genome paper}
Sporozoites were stored in 1 mL Trizol until RNA-isolation. 
Total RNA was isolated using the PureLink RNA Mini Kit (Invitrogen).

\subsection{Sequencing, quality assessment and alignment}
cDNA samples were sequenced by either GAIIX or Illumina Hiseq 2000 as specified in SI xx (both unstranded). Fastq\_quality\_filter was applied to Illumina Hiseq 2000 samples. 
A phred score of 10 was applied. We further set q = 60. These settings require that nine out of ten bases
or more are correct in at least 60\% of the bases for each read.

\subsubsection{Alignment and reference genomes}
We used the published \textit{Mus musculus} mm10 assembly (Genome Reference Consortium Mouse 
Build 38, GCA\_000001635.2) as reference genome including annotations for mouse data. The
\textit{E. falciformis} genome (Heitlinger14) was downloaded from ToxoDB (Gajria07). For the
alignment, the mouse and parasite genome files were merged into a dual reference genome, and 
files including mRNA sequences from both species were aligned against the dual reference genome
using TopHat2 (version 2.0.14, Trapnell09)/ Bowtie2 (version 1.1.2, Langmead12). Single-end and 
pair-end sequence samples were aligned separately with library type 'fr-unstranded' specified 
for pair-end samples. Import into R was enabled by the R package Ballgown, which requires bam 
files to be processed by Tablemaker (Frazee15). Tablemaker in turn makes use of Cufflinks 
(version 2.1.1, Trapnell10).

@@

\section*{Figures}
\begin{figure}[h!]
  \includegraphics[width=0.5\linewidth]{efmm-ratio}
  \caption{\csentence{Parasite read sequences per mouse read sequences for each experimental condition.}
  The more advanced the infection is, the higher the ratio of parasite reads is, reflecting parasite replication between day zero and day seven. ......compare 1st and 2nd ....... patterns between Rag and NMRI/C57BL/6?  ........ Values are mean of replicates (n=2, if * n=1) on log10 scale. Each sample (replicate) consists of mRNA from three different mice.}
  \end{figure}

  \begin{figure}[h!]
  \includegraphics[width=0.7\linewidth]{ef-heatmap}
  \caption{\csentence{Parasite genes with different mRNA abundance between samples, sorted into five clusters by hierarchical clustering (method = complete, distance = euclidean).}
      \textit{E. falciformis} samples from seven days p.i. cluster together (NMRI mice only). These samples have distinct mRNA abundance patterns in all gene clusters, although more pronounced in clusters 1 (up) and 4 (down). Distinct groups of genes also define sporozoites (cluster 5, up) and oocysts (cluster 2, up). mRNA profiles on days three and five p.i. from all three mouse strains cluster together. These samples are distinct from oocysts, NMRI day 7 p.i., and sporozoites, however closest to the latter. On scale bar, 0 is mean mRNA abundance for each gene. Up and downregulation is standard deviations from mean.}
      \end{figure}

\begin{figure}[h!]
  \includegraphics[width=0.7\linewidth]{mm-heatmap}
  \caption{\csentence{Mouse genes with different mRNA abundance between samples, sorted into four clusters by hierarchical clustering (method = complete, distance = euclidean).}
Three out of four samples from day 7 p.i. (NMRI only) cluster together. These samples are characterized by genes in cluster 2 (up), 3, and 4 (down). NMRI and C57BL/6 uninfected samples cluster together (left), defined by clusters 3, 4 (up), and 1 (down). Rag1-/- samples cluster together. These samples share a weak downregulation of most genes in cluster 2 as well as upregulation of a small group genes in the same cluster. Uninfected Rag1-/- samples are separated from infected ones with distinct profiles in clusters 1, 3, and 4. On scale bar, 0 is mean mRNA abundance for each gene. Up and downregulation is standard deviations from mean.}
      \end{figure}

>>>>

\usepackage[colorlinks]{hyperref} % To color blocks of text 
%\usepackage[dvipsnames]{xcolor} % More color options (see Wikibooks)  
\usepackage{changepage}
\usepackage{booktabs} % For prettier tables
\usepackage{float} % To control positions of e.g. tables or figures
\usepackage[section]{placeins}
\usepackage{graphicx}
\graphicspath{ {../} }
%\def\includegraphic{}
%\def\includegraphics{}

%%% Put your definitions there:
\startlocaldefs
\endlocaldefs

@@

\title{Dual host-parasite transcriptomes of apicomplexan Eimeria falciformis and its natural mouse host}

@@

\author[
   addressref={aff1},
]{\inits{TK}\fnm{Totta} \snm{Kasemo}}
\author[
   addressref={aff1},
]{\inits{SS}\fnm{Simone} \snm{Spork}}
\author[
   addressref={aff2},
]{\inits{CD}\fnm{Christoph} \snm{Dieterich}}
\author[
   addressref={aff1},
]{\inits{RL}\fnm{Richard} \snm{Lucius}}
\author[
   addressref={aff1, aff3},                   % id's of addresses, e.g. {aff1,aff2}
   corref={aff1},                       % id of corresponding address, if any
   noteref={n1},                        % id's of article notes, if any
   email={emanuel.heitlinger@hu-berlin.de}   % email address
]{\inits{EH}\fnm{Emanuel} \snm{Heitlinger}}

@@

\address[id=aff1]{%                           % unique id
  \orgname{Department of Biology, Humboldt-Universitat zu Berlin}, % university, etc
  \street{Philippstr. 13, Haus 14},                     %
  %\postcode{}                                % post or zip code
  \city{Berlin},                              % city
  \cny{Germany}                                    % country
}
\address[id=aff2]{%  
  \orgname{Max Planck Institute for Biology of Ageing},
  \street{Joseph-Stelzmann-Strasse 9B},
  \postcode{50931}
  \city{Cologne},
  \cny{Germany}
}
\address[id=aff3]{%  
  \orgname{Department Reproduction Biology and Evolutionary Ecology, Leibniz Institute of Zoo and Wildlife Research},
  \street{Alfred-Kowalke-Str. 17},
  \postcode{10315}
  \city{Berlin},
  \cny{Germany}
}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                          %%
%% Enter short notes here                   %%
%%                                          %%
%% Short notes will be after addresses      %%
%% on first page.                           %%
%%                                          %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%\begin{artnotes}
%\note{Sample of title note}     % note to the article
%\note[id=n1]{Equal contributor} % note, connected to author
%\end{artnotes}

%\end{fmbox}% comment this for two column layout

@@

\end{abstractbox}
%
\end{fmbox}% uncomment this for twcolumn layout

@@

  \section*{Results}
Text for this section \ldots

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% 	OVERVIEW READS PER SAMPLE, LAYOT AS EXPERIMENTAL OVERVIEW
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Experimental overview}
Replicate average of transcripts as order of magnitude for \textit{E. falciformis}/mouse are indicated in Table.... Columns in table represent mouse strains used in infection experiments. Rows represent timepoints post infection plus oocyst and sporozoite samples. The upper part of the table shows data for first infection, and oocyst and sporozoite data. The lower part shows data for challenge infection. Averages were calculated after sample exclusions (see Methods). For exact values, see Table 1.

	\setlength{\tabcolsep}{14pt}
	\begin{table}[H]
	\small
	\begin{center}
	\caption{Experimental overview with number of read for parasite/mouse indicated.}
\begin{tabular}{*4l}    \toprule
\textit{Day, 1st infection}  	& NMRI  & C57BL/6  & Rag1-/- \\ \midrule
	0 (control)    & $10^4$ / $10^8$  & $10^2$ / $10^8$  & $10^2$ / $10^7$  \\ %rep1/rep2, NA=no replicate
3  		& $10^5$ / $10^8$ & NA  & NA \\ 
5  		& $10^7$ / $10^8$ & $10^5$ / $10^7$  & $10^5$ / $10^7$ \\
7  		& $10^8$ / $10^7$ & NA  & NA \\ 
Oocysts 	& $10^8$ / NA & NA  & NA \\ 
Sporozoites 	& $10^7$ / NA & NA  & NA \\ \midrule

\textit{Day, 2nd infection}  	\\ \midrule
0 (control)     & $10^3$ / $10^8$  &  NA  & NA  \\ %rep1/rep2, NA=no replicate
3  		& $10^4$ / $10^8$ & NA  & NA \\ 
5  		& $10^4$ / $10^8$ & $10^5$ / $10^7$  & $10^5$ / $10^7$ \\
7  		& $10^6$ / $10^8$ & NA  & NA \\

	\bottomrule
 \hline
\end{tabular}
\end{center}
\end{table}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% 	NUMBER OF GENES DIFFERENT IN DIFFERENT COMPARISONS	
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\clearpage
\section{mRNA abundance differences between different experimental groups}
\setlength{\tabcolsep}{8pt}
\begin{table}[H]
\small
\begin{center}
\caption{mRNA abundance differences between different experimental groups.}
%\label{tab:table3}
\begin{tabular}{*3l}    \toprule
\textit{Day post infection} & \textit{Ef} genes different & Mouse genes different \\ 
	\textit{comparisons} 	    & (FDR$\leq$0.01) &  (FDR$\leq$0.01/0.05) \\ \midrule
	NMRI 0 vs NMRI 3		& NA   & 274 \\
	NMRI 0 vs NMRI 5		& NA   & 1736 \\
	NMRI 0 vs NMRI 7		& NA   & 2802 \\
	NMRI 3 vs NMRI 5     		& 111  & 1 \\
	NMRI 3 vs NMRI 7  		& 1385 & 1407 \\ 
	NMRI 5 vs NMRI 7  		& 1895 & 873 \\ 
	C57BL/6 0 vs C57BL/6 5		& NA	& 914 \\
	Rag1-/- 0 vs Rag1-/- 5		& NA	& 45 \\ \midrule
\textit{First and second infection} & 		 & 	 \\ 
\textit{comparisons} 	    & 		& 	\\ \midrule
	NMRI 3 1st vs NMRI 3 2nd  	& 0  & 5 \\
	NMRI 5 1st vs NMRI 5 2nd  	& 0  & 1 \\
	NMRI 7 1st vs NMRI 7 2nd  	& 0  & 902 \\
	C57BL/6 1st vs C57BL/6 2nd (day 5) & 0 &  mouse \\
	Rag1-/- 1st vs Rag1-/- 2nd (day 5) & 0 & mouse  \\ \midrule
\textit{Mouse strain comparisons} & 		 & 	 \\ \midrule
	NMRI vs C57BL/6 		& 22 	& NA* \\		
	NMRI vs Rag1-/-     		& 0 & NA* 	\\
	C57BL/6 vs Rag1-/-  		& 0   & 356 	\\ \midrule
\textit{Day post infection,} & 		 & 		 \\ 
\textit{parasite relevant comparisons} 	    & 		&  \\ \midrule
	Oocysts vs NMRI 3  	& 3310 & NA \\  
	Oocysts vs NMRI 5	& 3605 & NA \\ 
	Oocysts vs NMRI 7	& 3085 & NA \\ 
	Oocysts vs sporozoites  & 3421 & NA \\
	Sporozoites vs NMRI 3 	& 1663 & NA \\
	Sporozoites vs NMRI 5 	& 1605 & NA \\
	Sporozoites vs NMRI 7 	& 2473 & NA \\ \bottomrule
	\hline
\end{tabular}
\end{center}
\end{table}

\textit{E. falciformis} data overview of pairwise comparisons and number of genes with differently abundant mRNAs per comparison. NMRI followed by number indicates day post infection (e.g. NMRI3 = \textit{E. falciformis} genes from NMRI mouse day 3 post infection). Genes with Benjamini-Hochberg corrected p-values $\leq$0.01 as implemented in edgeR are indluced. NAs are missing samples or not applicable for the species. NA* is due to missing NMRI day 0 sample from first infection.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\clearpage
\subsection{Life cycle analysis... \textit{E. falciformis}}
\begin{figure}[h!]
%\begin{center}
%\centering
	\includegraphics[width=\linewidth]{EfLifecycleHeatmap.pdf}  
	\caption{\textit{E. falciformis} mRNAs with significantly different abundance at different times post infection in NMRI mice (see Table ...).  \textit{E. falciformis} samples from seven days p.i. cluster together (NMRI mice only). For \textit{E. falciformis} day 7, NMRI, the pattern is most pronounced in cluster 1 (up). Within this cluster, the second replicate from the challenge infection has a deviant profile (compare mouse data in Figure 6). Distinct groups of genes also define sporozoites (cluster 6: up) and oocysts (cluster 3: up, cluster 7: down). mRNA profiles on days three and five p.i. from all three mouse strains cluster together. These samples are distinct from oocysts, NMRI day 7 p.i., and sporozoites. The latter cluster closer with days 3 and 5 than with oocysts or day 7.}
%\end{center}
\end{figure}

\clearpage
\subsection{Parasite development in mouse data}
\begin{figure}[h!]
%\begin{center}
%\centering
\includegraphics[width=\linewidth]{MmLifecycleHeatmap.pdf}  
\caption{Mouse mRNAs with significantly different abundance at different times post infection in NMRI mice (see Table 3).
Three out of four samples from day 7 p.i. (NMRI only) cluster together. These samples are characterized by genes in cluster 2 (up) as well as 3 and 4 (down). The fourth day 7 sample (challenge infection, second replicate) clusters with day 3 and 5 samples but upon visual inspection of cluster 4 diplays a profile similar to non-infected mice. The same sample is abnormal in the parasite profile (see Figure 5). NMRI and C57BL/6 uninfected samples cluster together, defined by clusters 3, 4 (up), and 1 (down). Day 3 and 5 samples cluster together, with Rag1-/- forming a separate group and Rag1-/- non-infected most distant in this sample cluster. All non-infected samples share a high mRNA abundance in cluster 3.}
%\end{center}
\end{figure}

\clearpage
\subsection{First versus second infection, mouse}
\begin{figure}[h!]
%\begin{center}
\includegraphics[width=\linewidth]{Mm1st2ndHeatmap.pdf}  
	\caption{Mouse mRNAs with significantly different abundance between first infection and challenge infection (see Table 3). Samples do not cluster into first and second infection clusters. The two major sample cluster each contain four challenge infection samples. }
%\end{center}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\subsection*{GO term enrichments in heatmap clusters}
The annotations referred to here are inferred from orthologs in other \textit{Eimeria} spp. or
in \textit{T. gondii}

\subsubsection*{Data overview}
In total, xx reads were sequenced. For each sample, mouse read numbers were $10^7$ - $10^8$
Quality filters applied resulted in xx reads being removed from further analysis

@@

\subsection{RNA extraction}
Total RNA was isolated from infected epithelial cells, sporozoites and sporulated oocysts 
using Trizol according to the manufacturer’s protocol (Invitrogen). High quality  \emph{what is the meaning of 'high quality' here?}
RNA was used to produce an mRNA library using the Illumina’s TruSeq RNA Sample Preparation guide.
\emph{stolen from genome paper}
Sporozoites were stored in 1 mL Trizol until RNA-isolation. 
Total RNA was isolated using the PureLink RNA Mini Kit (Invitrogen) and reverse transcribed into cDNA.

\subsection{Sequencing, sequence quality assessment and alignment}
cDNA samples were sequenced by either GAIIX or Illumina Hiseq 2000 as specified in SI xx (both unstranded). A fastq\_quality\_filter (FASTQ-toolkit, version 0.0.14, available at https://github.com/agordon/fastx\_toolkit.git) was applied to Illumina Hiseq 2000 samples after replacing "N" bases by "." annotation.  
A phred score of 10 was applied. We further set q = 60. These settings require that nine out of ten bases
or more are correct in at least 60\% of the bases for each read.

\subsubsection{Alignment and reference genomes}
We used the published \textit{Mus musculus} mm10 assembly (Genome Reference Consortium Mouse 
Build 38, GCA\_000001635.2) as reference genome including annotations for mouse data. The
\textit{E. falciformis} genome (Heitlinger14) was downloaded from ToxoDB (Gajria07). For the
alignment, the mouse and parasite genome files were merged into a dual reference genome, and 
files including mRNA sequences from both species were aligned against the dual reference genome
using TopHat2 (version 2.0.14, Trapnell09) with -G specified, and a Bowtie2 (version 1.1.2, Langmead12) index of the dual genome. Single-end and pair-end sequence samples were aligned separately with library type 'fr-unstranded' specified for pair-end samples. Import into R was enabled by the R package Ballgown, which requires bam files to be processed by Tablemaker (Frazee15), in our case used with -qW -G specified. Tablemaker in turn makes use of Cufflinks (version 2.1.1, Trapnell10).

\subsubsection{Data normalisation and sample exclusions}
Count data was normalised in R, using the upperquartile method as implemented in edgeR (version 3.14.0). Briefly, normalisation assumes a negative binomial distribution of data and dispersed variance for genes with higher mean abundance. The normalisation shrinks the dispersion of variance towards a Poisson (linear) relationship between mean and variance. We evaluated the density distribution of our data and two samples contradicted the assumed negative binomial distribution and were excluded from further analysis: NMRI\_1st\_3dpi\_rep1 and NMRI\_2nd\_5dpi\_rep1 (SI ...).

\subsection{Differentially abundant mRNAs used for hierarchical clustering}
A selection of differently abundant mRNAs are used for hierarchical clustering of \textit{E. falciformis} life cycle relevant genes. In each comparison (see Table 3), the 500 differentially abundant mRNAs with lowest FDR are selected. In the next step, the 500 mRNAs from each comparison (or less) are joined (4935 genes) and only unique genes are selected (1618 genes). The 22 genes in the NMRI vs C57BL/6 comparison are not included in the \textit{E. falciformis} life cycle analysis.

All parasite samples were included in heatmap of genes detected as differentially abundant in timepoint comparisons.

On scale bar in heatmaps, 0 is mean mRNA abundance for each gene (row). Up (green) and down-regulation (brown) numbers denote number of standard deviations from row mean. Hierarchical clustering was performed using with Euclidean distances, method 'complete' (R package base).

\setlength{\tabcolsep}{10pt}
\begin{table}[H]
\small
\begin{center}
\caption{Differentially abundant \textit{E. falciformis} mRNAs used for hierarchical clustering.}
\begin{tabular}{*2l}    \toprule
	\textit{Data description} & Number of genes \\ \midrule
	Sum of 1st infection NMRI sample differences	& 4935  \\ 
	(including oocysts and sporozoites)	\\	
	Used in hierarchical clustering (heatmap)  	& 1618 \\ 	\bottomrule	
\hline
\end{tabular}
\end{center}
%\end{adjustwidth}
\end{table}

@@

\section*{Figures}
\begin{figure}[h!]
%  \includegraphics[width=0.5\linewidth]{efmm-ratio}
  \caption{\csentence{Parasite read sequences per mouse read sequences for each experimental condition.}
  The more advanced the infection is, the higher the ratio of parasite reads is, reflecting parasite replication between day zero and day seven. ......compare 1st and 2nd ....... patterns between Rag and NMRI/C57BL/6?  ........ Values are mean of replicates (n=2, if * n=1) on log10 scale. Each sample (replicate) consists of mRNA from three different mice.}
  \end{figure}

 % \begin{figure}[h!]
 % \includegraphics[width=0.7\linewidth]{ef-heatmap}
 % \caption{\csentence{Parasite genes with different mRNA abundance between samples, sorted into five clusters by hierarchical clustering (method = complete, distance = euclidean).}
 %     \textit{E. falciformis} samples from seven days p.i. cluster together (NMRI mice only). These samples have distinct mRNA abundance patterns in all gene clusters, although more pronounced in clusters 1 (up) and 4 (down). Distinct groups of genes also define sporozoites (cluster 5, up) and oocysts (cluster 2, up). mRNA profiles on days three and five p.i. from all three mouse strains cluster together. These samples are distinct from oocysts, NMRI day 7 p.i., and sporozoites, however closest to the latter. On scale bar, 0 is mean mRNA abundance for each gene. Up and downregulation is standard deviations from mean.}
 %     \end{figure}

%\begin{figure}[h!]
%  \includegraphics[width=0.7\linewidth]{mm-heatmap}
%  \caption{\csentence{Mouse genes with different mRNA abundance between samples, sorted into four clusters by hierarchical clustering (method = complete, distance = euclidean).}
%Three out of four samples from day 7 p.i. (NMRI only) cluster together. These samples are characterized by genes in cluster 2 (up), 3, and 4 (down). NMRI and C57BL/6 uninfected samples cluster together (left), defined by clusters 3, 4 (up), and 1 (down). Rag1-/- samples cluster together. These samples share a weak downregulation of most genes in cluster 2 as well as upregulation of a small group genes in the same cluster. Uninfected Rag1-/- samples are separated from infected ones with distinct profiles in clusters 1, 3, and 4. On scale bar, 0 is mean mRNA abundance for each gene. Up and downregulation is standard deviations from mean.}
%      \end{figure}
